Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments

### 2016-02-27 paper.tex
Supplement work.
Supplemental stuff about propcor.

% TODO: wtf does this break things?
%\externaldocument{supplement}

@@

\section{Results}

@@

Comparisons of gene or transcript expression are sensitive to how one chooses to
measure similarity. Because gene expression typically varies across many orders
of magnitude, measurements like the Pearson correlation, $L_1$ or $L_2$
distance, or root-mean-square error are often dominated by the most highly
expressed genes. Recognizing this, a popular alternative is Spearman's rank
correlation, which conversely can overemphasize the majority of transcripts,
which have no, or very low expression. These issues are discussed in depth by
Lovell et al \cite{Lovell:2015il}, who propose measurements of proportionality
as a more meaningful metric. We adopt their ``proportionality correlation'',
defined as $\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) /
(\text{var}(\log(x)) + \text{var}(\log(y)))$, where is $\text{cov}(x)$ and
$\text{var}(x)$ give the covariance and variance of $x$, respectively.  Similar
to other measures of correlation, proportionality correlation varies from -1 to
1, with 1 indicating perfectly proportionality, and -1 perfect reciprocality.
Zeros are accounted for by converting all measurements to transcripts per
million (TPM) and adding 0.1.
% TODO: add supplementary info about why 0.1 is a reasonable value.

@@

RSEM was used to produce both maximum likelihood and posterior mean estimates,
which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.  This is not an
exhaustive account of such methods, but does represent a wide variety of popular
approaches. When possible, the same read alignments were used for each method,
which notably excludes the alignment-free methods Sailfish, Salmon, and
Kallisto.
% TODO: see supplementary section XXX

@@

\begin{table}
\label{table:rlsim}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
\textbf{Isolator} & \textbf{0.919}
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{
Proportionality correlation between ground truth and estimates
produced by each method using simulated RNA-Seq reads.}
\end{table}

>>>>

\externaldocument{supplement}

@@

\section{Results}\label{results}

@@

Comparisons of gene or transcript expression are sensitive to how one chooses to
measure similarity. Because gene expression typically varies across many orders
of magnitude, measurements like the Pearson correlation, $L_1$ or $L_2$
distance, or root-mean-square error are often dominated by the most highly
expressed genes. Recognizing this, a popular alternative is Spearman's rank
correlation, which conversely can overemphasize the majority of transcripts,
which have no, or very low expression. These issues are discussed in depth by
Lovell et al \cite{Lovell:2015il}, who propose measurements of proportionality
as a more meaningful metric. We adopt their ``proportionality correlation'',
defined as $\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) /
(\text{var}(\log(x)) + \text{var}(\log(y)))$, where is $\text{cov}(x)$ and
$\text{var}(x)$ give the covariance and variance of $x$, respectively. Similar
to other measures of correlation, proportionality correlation varies from -1 to
1, with 1 indicating perfectly proportionality, and -1 perfect reciprocality.
Zeros are accounted for by converting all measurements to transcripts per
million (TPM) and adding 0.1. Results were found to be fairly insensitive to the
precise additive constant used here (Supplementary Section
\ref{additive-constant}). Though we believe proportionality correlation is a
superior approach, much of this analysis was also conducted using Spearman's
rank correlation in Supplementary Section \ref{spearman-reanalysis}, and largely
agrees with the results shown here.

@@

RSEM was used to produce both maximum likelihood and posterior mean estimates,
which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.  This is not an
exhaustive account of such methods, but does represent a wide variety of popular
approaches. When possible, the same read alignments were used for each method,
which notably excludes the alignment-free methods Sailfish, Salmon, and
Kallisto.
% TODO: see supplementary section XXX (how reads were aligned)

@@

\begin{table}
\label{table:rlsim}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & \textbf{0.919}
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{
Proportionality correlation between ground truth and estimates
produced by each method using simulated RNA-Seq reads.}
\end{table}

### 2016-02-27 supplement.tex
Supplemental stuff about propcor.

Existing statisticaly models (see {[}@Pachter:2011wm{]} for a review),
assume fragments are sampled uniformly at random from a transcript. This
does not exactly match the implications of random fragmentation in
which, under an ideal model, \emph{breakpoints} rather than fragments
are introduced uniformly at random. For a fragment to be observed it
must pass size selection and have fallen between two breakpoints, or one
breakpoint and the end of the transcript. Since the ends of a transcript
act as fixed breakpoints, the result isAsome enrichment of fragments at
either end of the transcript.

@@

\begin{figure}[H]
\includegraphics[width=0.2\textwidth]{analysis/cardio/colorkey.pdf} \\
\includegraphics[width=\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/IMMT-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
\end{figure}
\begin{figure}[H]
\ContinuedFloat
\includegraphics[width=\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/PALM-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
\caption{Point estimates of fold change of exon expression from qPCR and and
RNA-Seq using various methods. Exon expression for RNA-Seq is computed by
estimating transcript expression, then summing the expression of all transcripts
that share a parcitular exon.}
\label{fig:cardiofoldchange}
\end{figure}

@@

Benchmarks in this paper rely heavily on quantifying the agreement between
expression estimates and qPCR, spike-in controls, estimates from other samples,
or ground truth (in the case of simulations). As discussed in Section
\ref{results}, our methodology uses proportionality correlation as an
alternative to more common measures of correlation or aggregate error.

@@

This problem is compounded by the use of different estimators in the methods
being compared.  When maximum likelihood or, depending on the choice of prior,
maximum posterior estimates are computed, many expression estimates will be
zero. On the other hand, posterior mean estimates (as is computed in Isolator,
BitSeq, and optionally RSEM) are inherently non-zero.

In deciding how to deal with zeros, or the lack of zeros, it is easy to
artificially advantage one approach or the other. For example, removing zeros,
disadvantages methods that have no zeros, whereas quantifying error with
functions like median relative difference (as was used to argue the accuracy of
Kallisto \cite{Bray:2015uj}), will disadvantage methods that do not produce
zeros, or otherwise report small numbers.
\footnote{
To see this, consider that the absolute relative difference
between 0 and $\epsilon$ for an arbitrarily large or arbitrarily small but
positive $\epsilon$ is $\frac{\epsilon - 0}{(\epsilon + 0)/2} = 2$. Thus, MRD
considers ``absent versus barely present'' to be equivalent to ``absent versus
highly expressed'', an outcome that is at best, unintuitive.

More generally, the scale-invariance of relative difference means

%% TODO: which benchmark should we use to explore these problems? Maybe the
% SEQC transcript "consistency" benchmark? Regardless, it seems I didn't save
% any of this data locally so I have to spin up an instance.
% Fuck, I've paid about $600 in fees. I need to stop this thing.

$\frac{\alpha |x_1 - x_2|}{\alpha()/2} = 2$

}

This presents two plausible strategies to compensate: truncate small values to
zero, or add a small constant to every value. Both

\subsection{The effects of truncation}

%This paper argues that posterior mean estimates have some practical benefits.
%And reporting non-zero estimates in all cases is justifiable from a Bayesian
%...

\subsection{Effects of varying additive constant on computed proportionality
            correlation}

As our analysis relies on adding a seemingly arbitrary constant $c$ to
expression values before computing correlation, a natural question is how
sensitive our results our to its precise value. In short, we believe it is quite
insensitive, with results largely consistent when $c$ is varied across several
orders of magnitude.

%% TODO: plot

@@

\subsection{Select analyses repeated with Spearman's rank correlation}

@@

\caption{
Tables from Figure 1 recomputed using Spearman's rank correlation.
}
\end{figure*}

\begin{table}
\begin{center}
\label{table:rlsim:spearman}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator   & 0.724
\\\addlinespace
Kallisto   & 0.727
\\\addlinespace
Salmon     & \textbf{0.728}
\\\addlinespace
RSEM/ML    & 0.712
\\\addlinespace
Cufflinks  & 0.636
\\\addlinespace
eXpress    & 0.605
\\\addlinespace
Sailfish   & 0.648
\\\addlinespace
RSEM/PM    & 0.605
\\\addlinespace
BitSeq     & 0.595
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Table 1 recomputed using Spearman's rank correlation.}
\end{center}
\end{table}

>>>>

Existing statisticaly models (see \cite{Pachter:2011wm} for a review),
assume fragments are sampled uniformly at random from a transcript. This
does not exactly match the implications of random fragmentation in
which, under an ideal model, \emph{breakpoints} rather than fragments
are introduced uniformly at random. For a fragment to be observed it
must pass size selection and have fallen between two breakpoints, or one
breakpoint and the end of the transcript. Since the ends of a transcript
act as fixed breakpoints, the result isAsome enrichment of fragments at
either end of the transcript.

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.2\textwidth]{analysis/cardio/colorkey.pdf} \\
\includegraphics[width=0.9\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
\includegraphics[width=0.9\textwidth]{analysis/cardio/IMMT-comparison.pdf}
\end{figure}
\begin{figure}[H]
\ContinuedFloat
\centering
\includegraphics[width=0.9\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
\includegraphics[width=0.9\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
\includegraphics[width=0.9\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
\end{figure}
\begin{figure}[H]
\ContinuedFloat
\centering
\includegraphics[width=0.9\textwidth]{analysis/cardio/PALM-comparison.pdf}
\includegraphics[width=0.9\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
\caption{Point estimates of fold change of exon expression from qPCR and and
RNA-Seq using various methods. Exon expression for RNA-Seq is computed by
estimating transcript expression, then summing the expression of all transcripts
that share a parcitular exon.}
\label{fig:cardiofoldchange}
\end{figure}

@@

Benchmarks in this paper rely heavily on quantifying the agreement between
expression estimates and qPCR, spike-in controls, estimates from other samples,
or ground truth (in the case of simulations). As discussed in Section
\ref{results}, our methodology uses proportionality correlation
\cite{Lovell:2015il} as an alternative to more common measures of correlation or
aggregate error.

@@

This problem is not adressed by using median relative difference (MRD), as was used to
argue the accuracy of Kallisto \cite{Bray:2015uj}. Though it may seem robust due
to the use of a median, it does not account for the compositional nature of
RNA-Seq abundance estimates. A large error in one highly expressed transcript
can inflate the relative error in all other transcripts, thus driving up the
median. Furthermore, relative error can be unintuitive in the presense of
zero estimates. To see this, consider that the relative difference
between 0 and $\epsilon$ for an arbitrary positive $\epsilon$ is $\frac{\epsilon
- 0}{(\epsilon + 0)/2} = 2$. Thus, MRD unintuitively treats ``absent versus
barely present'' to be equivalent to ``absent versus highly expressed''.

For these reasons, we believe proportionality correlation is a more appropriate
measure when evaluating relative expression estimates.  The prevailing practical
issue with proportionality correlation is that it is defined only over positive
numbers. Methods that use maximum likelihood (or, depending on the choice of
priors, posterior maximum) estimates do report zeros.  We account for this by
adding a small constant ($c = 0.1 \text{TPM}$) to all expression values before
computing proportionality correlation.

\subsection{Effects of varying additive constant on computed proportionality
            correlation}\label{additive-constant}

As our analysis relies on adding a seemingly arbitrary constant $c$ to
expression values before computing correlation, a natural question is how
sensitive our results our to its precise value. In short, we believe it is quite
insensitive, with results largely consistent when $c$ is varied across several
orders of magnitude (Supplementary Figure \ref{fig:propcor-constant}).

\begin{figure*}
    \centering
    \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=0.8\textwidth]{pc-constant-seqc-primepcr.pdf}
    \caption{}
    \end{subfigure}

    \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=0.8\textwidth]{pc-constant-seqc-transcript-expression-consistency.pdf}
    \caption{}
    \end{subfigure}

    \caption{To account for zeros and near-zero values, a small constant $c$ is
        added to expression values before computing proportionality correlation.
        Throughout the paper we use $c = 0.1 \text{TPM}$. Here we varied the
        value of $c$ across several orders of magnitude and recomputed
        proportionality correlation in two of the benchmarks used in the paper:
        (a) comparison to PrimePCR qPCR and (b) consistency of transcript
        expression in SEQC data. Though the computed correlation changes
        depending on the value of $c$, methods largely maintain the same order.
    }
    \label{fig:propcor-constant}
\end{figure*}

@@

\subsection{Select analyses repeated with Spearman's rank correlation}\label{spearman-reanalysis}

@@

\caption{
Results from Figure 1 from the main paper recomputed using Spearman's rank correlation.
}
\end{figure*}

\begin{table}
\begin{center}
\label{table:rlsim:spearman}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator   & 0.724
\\\addlinespace
Kallisto   & 0.727
\\\addlinespace
Salmon     & \textbf{0.728}
\\\addlinespace
RSEM/ML    & 0.712
\\\addlinespace
Cufflinks  & 0.636
\\\addlinespace
eXpress    & 0.605
\\\addlinespace
Sailfish   & 0.648
\\\addlinespace
RSEM/PM    & 0.605
\\\addlinespace
BitSeq     & 0.595
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Table 1 from the main paper recomputed using Spearman's rank correlation.}
\end{center}
\end{table}

\bibliographystyle{plainnat}
\bibliography{references,extra-references}
